Further Evidence for Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution

### 2021-12-15 src/ms.tex
addressing co-author comments

Notably, both models fail to reproduce the observed short-period edge, possibly due to the assumption of solid-body rotation in both models. At early times, Sun-like and low-mass stars are expected to have strong radial differential rotation due to their rapid collapse onto the main-sequence. The core and envelope at these times are thus assumed to be decoupled. However, the core and envelope are expected to couple on timescales of a few tens of million years for Sun-like stars \citep{Denissenkov2010, GalletBouvier2015, Lanzafame2015} or hundreds of million years for low-mass stars \citep{GalletBouvier2015}. When the core and envelope couple, the star experiences a spin-up torque such that so-called ``two-zone'' models \citep{MacGregor1991} spin more rapidly than solid-body rotators at the same age. This hypothesis is discussed further in \S\ref{subsec:shortperiod}.

@@

Unlike the long-period pile-up, the short-period pile-up is not predicted by the WMB model. However, this feature may also be due to an epoch of stalled wind-driven angular momentum losses, albeit a temporary one. Works examining the rotation sequences of open clusters have found overlap between low-mass members of clusters of different ages \citep{Agueros2018, Curtis2019a, Curtis2020}. In other words, the spin rates of low-mass stars appear to evolve very little in the time elapsed between the ages of those clusters.

\citet{Curtis2019a, Curtis2020} proposed that such an effect could be produced by a temporary epoch of stalled spin-down, caused either by (a) core–envelope momentum transfer which offsets the effect of magnetic braking \citep[e.g.][]{MacGregor1991}, or (b) a reduction in the magnetic braking torque. Angular momentum is exchanged between the envelope and the core on a characteristic timescale known as the core–envelope coupling timescale. Theoretical predictions for this timescale are mass-dependent, and range from 30--110 Myr for a solar-mass star \citep{Bouvier2008, IrwinBouvier2009, Denissenkov2010, GalletBouvier2015, Spada2020}. In the core–envelope coupling scenario, wind-driven spin-down stalls when the star's age becomes comparable to the core–envelope coupling timescale. Stars then resume spinning down via magnetic braking, and from open cluster data \citet{Curtis2020} estimated that solar-mass stars resume spin-down after $\approx$~230~Myr. This timescale is consistent with independent studies of open clusters, which show multiple cluster sequences of Sun-like stars overlapping at ages of $\sim$100--200~Myr, with spin-down resuming by $\sim$300~Myr \citep{Fritzewski2020,Fritzewski2021}.

@@

We emphasize that the temporary epoch of stalled magnetic braking proposed by \citet{Curtis2019a, Curtis2020} is not to be confused with the termination of magnetic braking that characterizes the WMB model of \citet{vanSaders2016, vanSaders2019}. The physical mechanisms thought to be responsible for each of these proposed stages of rotational evolution are distinct. As mentioned above, one theory for the earlier stage of stalled braking is core–envelope coupling. Crucially, in the stalled braking phase, wind-driven angular momentum losses are not ceased but rather offset by the spin-up torque from core–envelope coupling. In contrast, in the WMB scenario, wind-driven angular momentum losses are proposed to cease ($dJ/dt$=0), with subsequent rotational evolution dictated by the changes in the moment of inertia. The physical mechanism responsible for WMB is not known, but one hypothesis is that the efficiency of magnetic braking declines with magnetic field complexity \citep[e.g.][]{Reville2015, vanSaders2016, Garraffo2016, Metcalfe2016, Metcalfe2019}.

@@

The short- and long-period pile-ups we examine here naturally produce a dearth of rotators at intermediate periods. This gap is the same period gap noticed by the authors mentioned above, as made apparent when comparing the \lamostmcq sample to the original \citet{McQuillan2014} sample. Moreover, as seen in Figure~\ref{fig:gap}, we recover the gradient in photometric variability across the gap pointed out by \citet{Reinhold2019}. The photometric variability, as measured through the \rper metric published by \citet{McQuillan2014}, reaches a local minimum near the location of the gap. This supports the notion that the cause of the gap is due to changes in the stars themselves, rather than being the result of mixed stellar populations. However, the variability levels on both sides of the gap are not close to the detection limit, as evidenced by the fact that periods are securely detected for stars with similar properties at much lower \rper values. This would suggest that the period gap is not due solely to a detection issue, unless variability levels were to drop precipitously as stars approached the gap.

@@

If the Sun is not in fact on the pile-up, one possibility is that the Sun evolved through a phase of weakened magnetic braking in the past and its subsequent period-lengthening was driven by its expansion away from the main sequence \citep[e.g.][]{vanSaders2019}. If this is the case, there is some moderate tension with the ages of pile-up stars in the CKS sample (Figure~\ref{fig:ages}). In the WMB model, hotter stars are expected to depart the pile-up on shorter timescales, yet some CKS stars in the pile-up appear to be both hotter and older than the Sun. It is of course possible that the isochrone ages are inaccurate, particularly for the cooler stars in the sample. If the Sun has indeed evolved through a period of weakened magnetic braking, this implies that the Sun's magnetic dynamo may be different from the majority of solar-type stars in the \citet{McQuillan2014} sample.

@@

    \item It is yet unclear whether the Sun resides on the long-period pile-up or has already evolved through it. Shifts of $\sim$50--300~K between the \teff scales of the Sun and the spectroscopic surveys studied here could place the Sun on the long-period pile-up. If the Sun has evolved through the pile-up, there is some modest tension with the Sun's age and the ages of the oldest stars on the pile-up, which might be explained by inaccurate isochrone ages.

@@

>>>>

Notably, both models fail to reproduce the observed short-period pile-up, possibly due to the assumption of solid-body rotation in both models. At early times, Sun-like and low-mass stars are expected to have strong radial differential rotation due to their rapid collapse onto the main-sequence. The core and envelope at these times are thus assumed to be decoupled. However, the core and envelope are expected to couple on timescales of a few tens of million years for Sun-like stars \citep{Denissenkov2010, GalletBouvier2015, Lanzafame2015} or hundreds of million years for low-mass stars \citep{GalletBouvier2015}. When the core and envelope couple, the star experiences a spin-up torque such that so-called ``two-zone'' models \citep{MacGregor1991} spin more rapidly than solid-body rotators at the same age. This hypothesis is discussed further in \S\ref{subsec:shortperiod}.

@@

Unlike the long-period pile-up, the short-period pile-up is not predicted by the WMB model. However, this feature may also be due to an epoch of stalled wind-driven angular momentum losses, albeit a temporary one as cluster studies demonstrate that stars continue to spin down beyond the short-period pile-up. Works examining the \teff–\prot sequences of open clusters have found overlap between low-mass members in clusters of different ages, notably Praesepe (0.67~Gyr), NGC 752 (1~Gyr), and NGC 6811 (1.4~Gyr) \citep{Agueros2018, Curtis2019a, Curtis2020}. In other words, the spin rates of low-mass stars appear to evolve very little in the time elapsed between the ages of those clusters. The short-period pile-up we observe may be the manifestation of the same type of slowed spin evolution, but for stars of higher masses and younger ages (since the pile-up more closely aligns with the Pleiades sequence) than the cluster members in the above-mentioned works.

\citet{Curtis2019a, Curtis2020} proposed that such an effect could be produced by a temporary epoch of stalled spin-down, caused either by (a)a reduction in the magnetic braking torque, or (b) core–envelope momentum transfer which offsets the effect of magnetic braking \citep[e.g.][]{MacGregor1991}. In the core–envelope momentum transfer scenario angular momentum is exchanged between the envelope and the core on a characteristic timescale known as the core–envelope coupling timescale. Theoretical predictions for the core–envelope coupling are mass-dependent, and for a solar-mass star range from 30--110 Myr \citep{Bouvier2008, IrwinBouvier2009, Denissenkov2010, GalletBouvier2015, Spada2020}. In the core–envelope coupling scenario, wind-driven spin-down stalls when the star's age becomes comparable to the core–envelope coupling timescale. Stars then resume spinning down via magnetic braking, and from open cluster data \citet{Curtis2020} estimated that solar-mass stars resume spin-down after $\approx$~230~Myr. This timescale is consistent with independent studies of open clusters, which show multiple cluster sequences of Sun-like stars overlapping at ages of $\sim$100--200~Myr, with spin-down resuming by $\sim$300~Myr \citep{Fritzewski2020,Fritzewski2021}.

@@

We emphasize that the temporary epoch of stalled magnetic braking proposed by \citet{Curtis2019a, Curtis2020} is not to be confused with the termination of magnetic braking that characterizes the WMB model of \citet{vanSaders2016, vanSaders2019}. The physical mechanisms thought to be responsible for each of these proposed stages of rotational evolution are distinct, though it is interesting that both the long- and short-period pileups seem to be well-described by curves of constant Rossby number. As mentioned above, one theory for the earlier stage of stalled braking is core–envelope coupling. Crucially, in the stalled braking phase, wind-driven angular momentum losses are not ceased but rather offset by the spin-up torque from core–envelope coupling. In contrast, in the WMB scenario, wind-driven angular momentum losses are proposed to cease ($dJ/dt$=0), with subsequent rotational evolution dictated by the changes in the moment of inertia. The physical mechanism responsible for WMB is not known, but one hypothesis is that the efficiency of magnetic braking declines with magnetic field complexity \citep[e.g.][]{Reville2015, vanSaders2016, Garraffo2016, Metcalfe2016, Metcalfe2019}.

@@

The short- and long-period pile-ups we examine here naturally produce a dearth of rotators at intermediate periods. This gap is the same period gap noticed by the authors mentioned above, as made apparent when comparing the \lamostmcq sample to the original \citet{McQuillan2014} sample. Moreover, as seen in Figure~\ref{fig:gap}, we recover the gradient in photometric variability across the gap pointed out by \citet{Reinhold2019}. The photometric variability, as measured through the \rper metric published by \citet{McQuillan2014}, reaches a local minimum near the location of the gap. This supports the notion that the cause of the gap is due to changes in the stars themselves, rather than being the result of mixed stellar populations. However, the variability levels on both sides of the gap are not close to the detection limit, as evidenced by the fact that periods are securely detected for stars with similar properties at much lower \rper values. This would suggest that the period gap is not due solely to a detection issue, unless variability levels were to drop precipitously as stars approached the gap. The data presented here provide evidence for the gap being caused by the same phenomena as the short-period pile-up, namely, core-envelope coupling. The decline in photometric variability that coincides with the gap is likely not the cause of it, but may result from a period of altered or decreased magnetic activity that is connected with core-envelope coupling.

@@

If the Sun is not in fact on the pile-up, one possibility is that the Sun evolved through a phase of weakened magnetic braking in the past and its subsequent period-lengthening was driven by its expansion away from the main sequence \citep[e.g.][]{vanSaders2019}. If this is the case, there is some moderate tension with the ages of pile-up stars in the CKS sample (Figure~\ref{fig:ages}). In the WMB model, hotter stars are expected to depart the pile-up on shorter timescales, yet some CKS stars in the pile-up appear to be both hotter and older than the Sun. It is of course possible that the isochrone ages are inaccurate, particularly for the cooler stars in the sample. If the Sun has indeed evolved through a period of WMB, this implies that the Sun may have a fundamentally different dynamo mechanism to the stars that have not yet transitioned through WMB. This suggests that the Sun's dynamo and consequent manifestations of magnetic activity may not be directly comparable or analogous to the majority of stars with measured Kepler rotation periods (as the majority have not yet transitioned through WMB). This may have implications for the validity of using the Sun as a benchmark calibrator and exemplar in stellar rotation and magnetism models.

@@

    \item It is yet unclear whether the Sun resides on the long-period pile-up or has already evolved through it. Shifts of $\sim$50--300~K between the \teff scales of the Sun and the spectroscopic surveys studied here could place the Sun on the long-period pile-up. If the Sun has evolved through the pile-up, there is some modest tension with the Sun's age and the ages of the oldest stars on the pile-up; some of the cooler long-period pile-up stars in the CKS sample are both more massive and older than the Sun, which contradicts the expectation from the WMB model that hotter stars spend a shorter period of time on the pile-up. However, this tension might be simply explained by inaccurate isochrone ages.

@@

    \item An increasing number of open clusters are being discovered by searching for a clustering of rotation periods along a slow-rotator sequence. (I think maybe Pisces Eridani is an example of this? Or perhaps Jason has a better example?). However, the long-period pile up discovered here can mimic a slow-rotator sequence in a small sample of unassociated stars with different ages and precisely measured temperatures and rotation periods. This is clearly demonstrated by the CKS sample in the right-hand panel of figure 1 -- taken out of context, these data look like a group of coeval stars with a slow-rotator sequence. The discovery of pile-ups in stellar rotation periods therefore has consequences for open cluster studies. When over-densities or ridges appear present in the rotation period distributions of stellar populations, care must be taken to ensure that these are not caused by WMB or core-envelope coupling before assuming that population is coeval and using the any over-densities to age-date it.
